Statistical analysis of whole-genome bisulfite-sequencing data

### 2014-12-03 Rmarkdown/wgbs_statistical_analysis.Rmd
Re-jigged DMRs and added PMD section.

- Chapter overview
  - One sample
    - Pre-sequencing
    - Post-sequencing
    - Some complications for a single sample
      - What is $\mathcal{I}$?
      - Genetic heterozygosity at a methylation locus
  - Multiple samples
    - Some complications with multiple samples
      - What is $\mathcal{I}$?
  - Parameter estimation
    - Estimating $M$, $U$
    - Estimating $B$
      - Beta-binomial models
      - Smoothing
    - Statistical properties of $\beta$-values
      - Empirical distributions
      * Write a function(s) to do the following and apply it to the EPISCOPE, Lister, Seisenberger and Ziller datasets:
        - Genome-wide distribution of CpG $\beta$-values
        * Kernel density estimates
        * Stratify by CGI status, genic, intragenic, repeats
        * Correlation across strands
        * Effect of sequencing coverage threshold (`min_cov`)
      - What is the genome-wide methylation level?
        * Various figures get bandied about; what's a good summary?
        - A criticism of ``identifying methylcytosines''
      - Correlations
      - Biases
      - Transformations
  - Downstream analyses
    - Differential methylation
      - Experimental design considerations
      - DMCs: Identify $\bar{B_{i}^{(1)}} \neq \bar{B_{i}^{(2)}}$
        - One-group experiments
        - Two-group experiments
        - Multi-group experiments
      - DMRs: Identify runs of DMCs
        - A priori vs. data-driven regions
        - One-group experiments
        - Two-group experiments
        - Multi-group experiments
    - Differential variability
      - DVCs: Identify $\sigma(B_{i}^{(1)}) \neq \sigma(B_{i}^{(2)})$
      - DVRs: Identify runs of DVCs
    - Allele-specific methylation
    - Methylation entropy
    - Epipolymorphism
    - Epiallele detection
    - Integrating DNA methylation data with other genomes data

@@

A common downstream analysis is the identification of differential methylation, for which many methods have been proposed \citep[e.g.,]{Akalin:2012cmChen:2013eh, Chen:2014jb, Dolzhenko:2014bo, Gokhman:2014cp, Jaffe:2012gx, Lacey:2013iy, Lister:2009hy, Rijlaarsdam:2014hp, Sofer:2013bk, Xie:2014ez, Feng:2014iq, Hebestreit:2013ko, Sun:2014fk, Park:2014ho, Hansen:2012gr}. This includes the identification of differentially methylated cytosines (DMCs) and differentially methylated regions (DMRs).

@@

Regardless of the test used, all attempts to identify DMCs from whole-genome bisulfite-sequencing data must pay a large multiple-hypothesis testing penalty. Correcting for multiple hypothesis testing is standard practice in the analysis of genomics data. but the sheer number of tests, in this case approximately 25 million, is at least an order of magnitude greater than commonly tested in other genomics experiments (e.g., approximately 1 million tests in genome-wide association studies and 20,000 tests in studies of differential gene expression). Various methods have been used to correct for this multiple testing, as illustrated in Table \ref{tab:multiple_testing}.

@@

\paragraph{One-group experiments}\label{sec:one_group_dmcs}

@@

\paragraph{Multi-group experiments}\label{sec:two_group_dmcs}

@@

Several software packages are now available for identifying DMCs, although some are limited to the two-group experiment. Rather than directly analysing the raw $\beta$-values, these software generally make additional modelling assumptions, such as the beta-binomial model (Section \ref{sec:beta-binomial_model}), and/or perform some transformation, such as smoothing of the $\beta$-values (Section \ref{transformation}).

@@

Not all methods for identifying differential methylation are designed for identifying DMCs. Both `bsseq` \citep{Hansen:2012gr} and `Aclust` \citep{Sofer:2013bk} are methods explicitly designed for identifying DMRs rather than DMCs and, in the case of `bsseq`.

@@

The idea of testing for differential methylation is relatively straightforward. For each sample, the region-wise methylation level is some summary of the methylation level at each loci in that region, for example, the weighted average of $\beta$-values for all loci in the region, where the weights are proportional to the sequencing coverage. These averages are then compared between samples. I describe basic implementations of this for both one-group and multi-group experiments, and then detail a more sophisticated analysis implemented in `BiSeq` \citep{Hebestreit:2013ko}.

@@

\subsubsection{Multi-group experiments}\label{sec:multi_group_dmrs}

__TODO__

\subsubsection{A more complex example}\label{sec:a_more_complex_example}

@@

The most common type of algorithm for identifying data-driven DMRs, and arguably also the most _ad hoc_, is based on detecting clusters of DMCs \citep[e.g.,]{Lister:2009hy, Lister:2011kg, Hansen:2011gu, Feng:2014iq}. The initial identification of DMCs will typically use a relaxed statistical significance threshold. For example, \cite{Hansen:2011gu} use all CpGs with a P-value in the lowest $5\%$ and declare putative DMRs to be contiguous runs of such CpGs that are within a given distance of one another and with "all differences in the same direction". This latter requirement is what I call the _consistency_ of the DMR (see Section \ref{sec:consistency_of_dmrs}).

@@

\subsection{PMDs}

- Lister
- Berman

@@

- Belinda
- Jaffe

>>>>

@@

A common downstream analysis is the identification of differential methylation. This includes the identification of differentially methylated cytosines (DMCs) and differentially methylated regions (DMRs). Many methods have been proposed \citep[e.g.,]{Akalin:2012cmChen:2013eh, Chen:2014jb, Dolzhenko:2014bo, Gokhman:2014cp, Jaffe:2012gx, Lacey:2013iy, Lister:2009hy, Rijlaarsdam:2014hp, Sofer:2013bk, Xie:2014ez, Feng:2014iq, Hebestreit:2013ko, Sun:2014fk, Park:2014ho, Hansen:2012gr}. All of these methods are designed for two-group experiments, with some being general enough to extend to multiple-group experiments[^multi-group_extension]. I describe in greater detail methods for identifying differential methylation in a one-group experiment, admitedly an unusual design but one that I faced during my PhD.

[^multi-group_extension]: The extension to multiple-group experiments isn't always implemented in the software, even if discussed in the paper.

@@

Regardless of the test used, all attempts to identify DMCs from whole-genome bisulfite-sequencing data must pay a large multiple-hypothesis testing penalty. Correcting for multiple hypothesis testing is standard practice in the analysis of genomics data, but the sheer number of tests, in this case approximately 25 million, is at least an order of magnitude greater than commonly tested in other genomics experiments[^multiple_testing]. Various methods have been used to correct for this multiple testing, as illustrated in Table \ref{tab:multiple_testing}.

[^multiple_testing]; For example, there are approximately 20,000 tests in studies of differential gene expression and one million tests in genome-wide association studies.

@@

\subsubsection{One-group experiments}\label{sec:one_group_dmcs}

@@

\subsubsection{Multi-group experiments}\label{sec:multi_group_dmcs}

@@

Several software packages are now available for identifying DMCs, although some are limited to the two-group experiment. Rather than directly analysing the raw $\beta$-values, these software generally make additional modelling assumptions, such as the beta-binomial model (see Section \ref{sec:beta-binomial_model}), and/or perform some transformation of the data, such as smoothing of the $\beta$-values (see Section \ref{transformation}).

@@

Not all methods for identifying differential methylation are designed for identifying DMCs. Both `bsseq` \citep{Hansen:2012gr} and `Aclust` \citep{Sofer:2013bk} are methods explicitly designed for identifying DMRs rather than DMCs.

@@

The idea of testing for differential methylation is relatively straightforward. For each sample, the region-wise methylation level is some summary of the methylation level at each loci in that region, for example, the weighted average of $\beta$-values for all loci in the region, where the weights are proportional to the sequencing coverage. These averages are then compared between samples. I describe a basic implementation of this for a one-group experiment and then detail a more sophisticated method that is implemented in `BiSeq` \citep{Hebestreit:2013ko}.

@@

\subsubsection{A more sophisticated multi-group analysis}\label{sec:a_more_complex_example}

@@

The most common type of algorithm for identifying data-driven DMRs, also arguably the most _ad hoc_, is based on detecting clusters of DMCs \citep[e.g.,]{Lister:2009hy, Lister:2011kg, Hansen:2011gu, Feng:2014iq}. The initial identification of DMCs will typically use a relaxed statistical significance threshold. For example, \cite{Hansen:2011gu} use all CpGs with a P-value in the lowest $5\%$ and declare putative DMRs to be contiguous runs of such CpGs that are within a given distance of one another and with "all differences in the same direction". This latter requirement is what I call the _consistency_ of the DMR (see Section \ref{sec:consistency_of_dmrs}).

@@

There are many other questions that can be addressed using bisulfite-sequencing data. Like differential methylation, some of these might be also be tackled with other methylation assays, such as microarrays. However, as we will see, some can only be answered by assays that have single nucleotide resolution, which is where bisulfite-sequencing can prove particularly powerful.

\subsection{Partially methylated domains}

Partially methylated domains, PMDs, are long stretches of the genome where the average methylation level, $B_{i}$, is away from the extremes of $0$ and $1$, typically in the range $0.20 - 0.70$. An example of such a region is shown in __FIGURE__ (__TODO: Find a PMD and plot it__).

PMDs were first identified in the IMR90 genome by \citet{Lister:2009hy}. Using a simple sliding window algorithm, \citeauthors{Lister:2009hy} unexpectedly found that approximately $40\%$ of every autosome was a PMD and that the average length of these PMDs was a large 153 kb. They also showed that these PMDs were not simply due to a methylated population and an unmethylated subpopulation of cells in the sample. This did this by showing that inidividual reads mapped to these PMDs contained both methylated and unmethylated bases.

A subsequent study found that PMDs are a common feature of somatic cell lines and that they comprise $> 30\%$ of the genome \citep{Lister:2011kg}. Perhaps even more intriguingly, across four somatic cell lines profiled with whole-genome bisulfite-sequencing, \citeauthor{Lister:2011kg} found a large amount of these genomes ($664$ Mb) comprised shared PMDs.

PMDs have also been identified within tumour methylomes \citep{Berman:2012ga, Hansen:2011gu} and \citet{Hansen:2011gu} found that these overlap with other important genomic features called large organized chromatin lysine modifications (_LOCKs_) and lamina associated domains (_LADs_). However, PMDs are conspicuous by their absence in pluripotent cell lines, including both embryonic stem cells and induced pluripotent stem cells \citep{Lister:2011kg}. Their absence in the induced pluripotent cell lines may reflect the fact that the methylome is 'reset' upon induction of pluripotency \citep{Lister:2011kg}.

More sophisticated methods have also been proposed to identify these PMDs. `methylSeekR` \citep{Burger:2013kq} is one such algorithm. It uses a hidden Markov model of the $\beta$-values to segment the genome into unmethylated, lowly methylated and partially methylayed regions. While `methylSeekR` is based on $\beta$-values and does not make use of individual reads to identify PMDs, it does require the high spatial resolution of $\beta$-values that methylation microarrays cannot provide.

@@

- Belinda
- Jaffe
- DVCs: Identify $\sigma(B_{i}^{(1)}) \neq \sigma(B_{i}^{(2)})$
- DVRs: Identify runs of DVCs
